Hysteresis in human computation:\\ how one task affects another

### 2015-01-08 docs/publications/2014.09_SCIENCE/man.tex
proofed latest draft; still need to update figure legends

Faced with the ever-growing scale and complexity of data, academic and 
industrial researchers are increasingly relying on \textit{human computation} 
as a means to extract useful knowledge.  Microtask
platforms, like Amazon Mechanical Turk, act like a human compute cluster,
making it possible to access human cognition quickly and inexpensively. 
This has lead to their broad adoption by social and computer scientists.  
Human computation is being used in applications
where expert opinion would normally be sought, such as in reviewing 
medical images.
Here we show that the repetition of similar tasks, which is an inherent
feature of microtask and crowdsourcing platforms, acts as a strong
and largely unexplored source of bias.  The effects of previous tasks on
later ones---\textit{intertask effects}---are more severe than the effects 
of framing, such as by disclosing a task's purpose.  
Using image-labeling, the canonical microtask, we show that prior 
tasks influence the content that workers chose to focus on, as well as the 
richness and specialization of their vocabulary.    
While intertask effects can be a source of systematic bias, our
results suggest that, when properly controlled, they might be leveraged
to hone worker focus and specificity, and could play a key role in engineering
tasks that elicit reproducible, expert-level judgments.
\end{sciabstract}

@@

Microtasks are typically distributed to workers in bundles.  For example, a 
worker might be assigned a set of ten images, and asked to provide 
descriptive words (or \textit{labels}) for each.  
On MTurk, these bundles are called Human Intelligence Tasks (HITs).
But for the sake of clarity, we will use ``task'' to mean the smallest 
repeatable unit of work, which, in our study, consists of labeling a single
image.

Results in psychology show that people are susceptible to various kinds of 
priming \cite{BJOP1796,No2007,beller1971priming}, including 
task-repetition effects \cite{Gass1999549,sohn2001task}.  
Yet to our knowledge, no study has investigated the bias that might
result from the fact that microtask workers typically perform many tasks in 
succession.  The experiments we present here show that a worker's response 
to a task depends strongly on the tasks she had performed beforehand.
In other words, unlike CPUs and GPUs, HPUs are subject to 
\textit{hysteresis}, meaning that their outputs depend on the 
\textit{history} of their inputs.
We call the effects that prior tasks induce on later ones 
\textit{intertask effects}.

We investigated intertask effects using image-labeling tasks on the Amazon 
Mechanical Turk (MTurk) microtask platform.  Image-labeling tasks are among
the most common MTurk tasks \cite{chandler2013breaking,Berinsky2012351,Finnerty2013,paolacci2010running}.  This trend will probably continue 
in the long term because of their role in computer vision research 
\td{find another application} \cite{5543192}.  In our first experiment, 
which we will refer to as 
\textit{intertask-food-objects}, workers were assigned to either a
\textit{food} or \textit{objects} treatment.  Workers in these treatments
performed a set of five \textit{initial tasks}, which each involved 
providing five descriptive labels for an image depicting 
either food or (non-food) objects, depending on the worker's 
treatment (see Fig.~\ref{fig:task}A and B).  Following the initial 
tasks, workers performed five \textit{test tasks}, which were identical for 
both treatments.  The test tasks contained images depicting a 
mixture of food and other objects (see Fig.~\ref{fig:task}C), and workers
were again asked to supply five labels.  Five initial 
tasks, together with the test tasks, comprised one HIT 
(\# workers were only allowed to participate once). During the HIT, workers 
were presented images one at a time, and, crucially, there was no distinction
or interruption between the initial and test tasks.

@@

While this affirms the \textit{existence} of intertask effects, we wish to
understand the nature and severity of the bias induced by prior task 
exposures.  
We define the extent of bias, $\theta$, induced by alternative exposures, to 
be the resulting change in the probability with which workers provide given 
responses (formally the totatal variation distance):
\begin{equation}
	\theta = \frac{1}{2}\sum_{x \in X} \left| p_1(x) - p_2(x) \right|,
\end{equation}
where $p_i(x)$ is the probability that a worker subjected to treatment $i$
provides response $x$, from the set of possible responses $X$.
The problem of determining $\theta$ from finite samples of responses is 
difficult, which leads us to reformulate it as a classification problem: 
given a worker's 
response, can a machine reliably infer what her priming exposure had been?  
Intuitively, the greater the effect that priming exposures have on workers' 
responses, the more accurate the classifier can be.  Formally, as we show
in the supplementary text, the performance of such a classifier, $\eta$,
yields a lower bound on the true bias:
\begin{equation}
	\theta \geq 2\eta - 1.
	\label{l1}
\end{equation}
Thus, measuring the performance of a classifier allows us to establish a
lower bound on the extent of bias. This approach can be used to bound the 
bias induced by any perturbation, including both intertask effects and 
framing.  As we have defined it, bias ranges from 0 (meaning the exposures 
do not affect the probabilities of respnoses), to 1 
(or 100\%) (meaning that under different exposures, workers provide completely
different sets of responses).

@@

Using a na\"ive Bayes classifier (\# in the 
supplementary material we discuss the choice of classifier, the
approach to measuring accuracy, and draw a comparison to the results obtained 
if a support vector machine is used.), we found that intertask effects 
lead to a 30\% bias between workers from the \textit{food} and 
\textit{objects} treatments.

As mentioned, the effects of framing on workers' responses to microtasks has 
been studied in various ways 
\cite{Kinnaird2012281,chandler2013breaking,thibodeau2013natural}.
Therefore, as a point of comparison, we performed a similar experiment 
in which we framed the purpose of the tasks.  Workers were either told that 
the tasks were ``Funded by the laboratory for the visual perception of Food 
and Ingredients'', or ``\ldots of Objects and Tools''.  We did not observe a 
statistically significant framing-induced bias, using a na\"ive Bayes 
classifier (see Fig.~\ref{fig:theta}).  This suggests
that prior task exposure is a stronger priming modality than framing.
\td{I need to talk about the $\chi^2$ result here.}

We conducted variants of these experiments using different images, to see 
whether this trend was robust.  In the experiment 
\textit{intertask-food-culture},
workers were either assigned to a \textit{food} or \textit{culture} treatment.
As before, the initial tasks for the \textit{food} treatment contained images 
depicting food (\#) (see Fig.~\ref{fig:task}D). 
In the \textit{culture} treatment, 
the initial tasks contained images depicting cultural scenes 
(of dance, sport, or music) (see Fig.~\ref{fig:task}E).  The test tasks, which were identical for both
treatments, included images depicting meals of identifiable cultural 
origin 
(see Fig.~\ref{fig:task}F).  Results for this experiment again showed a 
strong bias as a result of intertask effects (about 50\%) 
(see Fig.~\ref{fig:theta}A).  Again, we compared intertask effects to 
framing, by performing an experiment (\textit{frame-food-culture}) using 
the same test tasks as in \textit{intertask-food-culture}, in which we exposed
workers to a statement framing the purpose of the tasks as the recognition 
of either food or cultures.  
For this framing experiment, we did not observe a statistically significant 
difference between the labels attributed by the \textit{food} and 
\textit{culture} treatments during test tasks (see Fig.~\ref{fig:theta}).

@@

To deepen our understanding, we investigated workers' lexical richness in 
reference to food, that is, the number of \textit{unique} food-related words
used.
Both \textit{intertask} experiments showed that food-exposed workers had 
greater lexical richness, in reference to food, than their non-food-primed 
counterparts (as much as 20\% more; see Fig.~\ref{fig:specificity}).  This
held in the case of \textit{intertask-food-culture}, even though, in that 
experiment, food-primed workers referred to food less often overall.  
This enrichment of the food-related lexicon 
was also observed in the \textit{priming-food-objects}, and 
\textit{echo-food-objects} experiments, but to a lesser extent.

This lexical enrichment suggests that priming, whether by prior task 
exposure or framing, leads workers to use more refined 
or specialized words, in reference to concepts present in the prime.
To test this directly, we used the wordnet corpus to operationalize the 
notion of word specialization.
Wordnet provides a set of hypernym-hyponym relations between \# English 
nouns.  
Hypernyms are generalizations (for example, ``bread'' is a hypernym for 
``pumpernickel''), while hyponyms are specializations.
To assess the relative specialization of words from two different treatments,
we compared all pairs of words from one treatment to the other. 
We tallied the cases where one treatment's word was more specialized
than the other, as well as the cases wher roles were reversed.  
The difference of these tallies, as a percentage of their total, served as 
the relative specialization of one treatment's vocabulary relative to the 
other (\#) (see Fig.~\ref{fig:specificity}).  
Using this approach, we compared the specialization of food-related words 
for the treatments of all experiments.

In all experiments except \textit{frame-food-culture}, food-primed workers 
used significantly more specialized words than their non-food-primed 
counterparts (about 15\% more; see Fig.~\ref{fig:specificity}C).
It is interesting that such substantial increases in both the lexical 
richness and specialization of food-related words 
held for \textit{intertask-food-culture}, where, again, we observed fewer 
references to food by food-primed workers overall. These observations point 
to countervailing factors: one factor tending to activate the more 
specialized and less common food-related words 
(yielding greater lexical richness and specialization), and the other tending 
to suppress certain, presumably more common or generic words 
(yielding fewer food-related words in total).

@@

Taken together, our results can be 
explained through a combination of positive and negative priming.
Positive priming (usually simply ``priming'') occurs when a prior stimulus 
predisposes a person to give certain responses in an ensuing task, and
is often observed as an increase in the speed or accuracy of a response, or
the ability to recognize features of a briefer, quieter, or noisier stimulus 
(\#).
Negative priming occurs when, after being repeatedly exposed to a stimulus 
considered to be non-salient, a person begins to ignore that 
stimulus (\#).

@@

Prior to this study, little thought appears to have been given to the 
ordering and bundling of microtasks.  But our findings show that these are 
powerful design factors.  A natural response to our findings 
might be simply to randomize task ordering, a practice that is commonly 
observed.  But the sheer extent of bias that we observed leads us to question 
whether that is the best approach, since this would tend to raise 
the ``noise floor'' of the study.

Furthermore, our findings suggest that, through careful task engineering,
it might be possible to achieve much more.  The boost that we observed to the 
richness and specialization of workers' responses motivates further 
investigation to acquire a deeper understanding of intertask effects.  
We hope that future work will yield techniques to leverage intertask effects, 
enabling one to tune the focus, diversity, and specificity of worker 
responses.

A consistent goal in human computation is the achievement of expert-level
judgments from non-expert workers (\#).  This has been achieved in some
applications (\#). The distinction between experts and novices can be 
attributed in part to the posession of specialized knowledge and heuristics. 
But to a large degree, this distinction lies in the ability of 
experts to direct their focus toward salient features, while filtering out 
non-salient ones \cite{kellman2009perceptual}.  Using prior task exposure to 
guide focus and salience attribution might enable expert-level 
judgment in a wider variety of crowdsourced applications.

Here we have shown that a large bias can be introduced into microtask 
responses, simply from the performance of earlier tasks. 
These effects are much stronger than framing.  
While our findings raise serious concerns about the 
current state of microtask design, they also reveal potential opportunities 
for more 
refined control over worker focus and acuity.  Any application that relies
on the repetitive use of human judgment is likely subject to the phenomena
we described here.  It remains to be determined by future work whether 
intertask effects can be leveraged for more performant and reliable human 
computation.

\bibliography{newbib}
\bibliographystyle{Science}

>>>>

Faced with the ever-growing scale and complexity of data, academic and 
industrial researchers are increasingly relying on \textit{human computation} 
to extract useful knowledge.  Microtask
platforms, like Amazon Mechanical Turk, act like a human compute cluster,
making it possible to access human cognition quickly and inexpensively. 
This has lead to their broad adoption by social and computer scientists.  
Human computation is being used experimentally in applications
where expert opinion would normally be sought, such as in reviewing 
medical images.
Here we show that the repeated performance of similar tasks, an inherent
feature of microtask and crowdsourcing platforms, acts as a strong
and largely unexplored source of bias.  Previous tasks exert strong 
influences on how workers perform later tasks, which we call 
\textit{intertask effects}.  These effects are more severe than effects 
caused by framing, such as by disclosing a task's purpose.  
Using image-labeling, the canonical microtask, we show that prior 
tasks influence the content that workers chose to focus on, as well as the 
richness and specialization of their vocabulary.    
While intertask effects can be a source of systematic bias, our
results suggest that with appropriate task-engineering, it might be possible
to leverage intertask effects to hone worker focus and specificity, helping
to elicit reproducible, expert-level judgments.
\end{sciabstract}

@@

Here we focus on an aspect of the microtask setup that is ubiquitous, but 
has not yet received much attention.  Microtasks are typically distributed 
to workers in bundles.  For example, a worker might be assigned a set of ten 
images, and asked to provide descriptive words (or \textit{labels}) for 
each.  On MTurk, these bundles are called Human Intelligence Tasks (HITs).
But for the sake of clarity, we will use ``task'' to mean the smallest 
repeatable unit of work, which, in our study, consists of labeling a single
image.

Results in psychology show that people are 
susceptible to various kinds of priming 
\cite{BJOP1796,No2007,beller1971priming}, including 
task-repetition effects \cite{Gass1999549,sohn2001task}.  
But to our knowledge, no study has investigated the bias that might
result from the fact that microtask workers typically perform many tasks in 
succession.  The experiments we present here show that a worker's response 
during a task depends strongly on the tasks performed beforehand.
In other words, unlike CPUs and GPUs, HPUs are subject to 
\textit{hysteresis}, meaning that their outputs depend on the 
\textit{history} of their inputs.
We call the effects that prior tasks induce on later ones 
\textit{intertask effects}.

We investigated intertask effects using image-labeling tasks on the Amazon 
Mechanical Turk (MTurk) microtask platform.  Image-labeling tasks are among
the most common MTurk tasks \cite{chandler2013breaking,Berinsky2012351,Finnerty2013,paolacci2010running}.  This trend will probably continue 
in the long term, because image labeling is still very difficult to automate
\cite{5543192}, and is used in many applications, such as website moderation, 
computer vision research, and social media analytics (\# insert other refs) 
\cite{5543192}.  In our first experiment, which we will refer to as 
\textit{intertask-food-objects}, workers were assigned to either a
\textit{food} or \textit{objects} treatment.  Workers in these treatments
performed a set of five \textit{initial tasks}, which each involved 
providing five descriptive labels for an image depicting 
either food or (non-food) objects, depending on the worker's 
treatment (see Fig.~\ref{fig:task}A and B).  Following the initial 
tasks, workers performed five \textit{test tasks}, which were identical for 
both treatments.  The test tasks contained images depicting a 
mixture of food and objects (see Fig.~\ref{fig:task}C), and workers
were again asked to provide five labels.  Five initial 
tasks, together with the test tasks, comprised one HIT 
(\# workers were only allowed to participate once). During the HIT, workers 
were presented images one at a time, and, crucially, there was no distinction
or interruption between the initial and test tasks.

@@

While this affirms the \textit{existence} of intertask effects, we wish to
understand the nature and severity of the bias induced by prior task 
exposures.  
We define the extent of bias, $\theta$, induced by alternative exposures, to 
be the resulting change in the probability with which workers provide given 
responses (formally the total variation distance):
\begin{equation}
	\theta = \frac{1}{2}\sum_{x \in X} \left| p_1(x) - p_2(x) \right|,
\end{equation}
where $p_i(x)$ is the probability that a worker subjected to treatment $i$
provides response $x$, from the set of possible responses $X$.
It is important to emphasize that a sigificant result from a $\chi^2$ test 
only allows us to conclude that $\theta$ is non-zero.  Determining $\theta$ from finite samples of 
responses, using purely statistical approaches, is difficult (\#), which 
leads us to reformulate the
problem in terms of classification: given a worker's 
response, can a machine reliably infer what her prior task-exposure had 
been?  Intuitively, the greater the effect that prior tasks have on subsequent
responses, the more accurate the classifier can be.  Formally, as we show
in the supplementary text, the accuracy of such a classifier, $\eta$,
yields a lower bound on the bias:
\begin{equation}
	\theta \geq 2\eta - 1.
	\label{l1}
\end{equation}
Thus, measuring the accuracy of a classifier allows us to establish a
lower bound on the extent of bias. This approach can be used to bound the 
bias induced by any perturbation, including both intertask effects and 
framing.  As we have defined it, bias ranges from 0 (meaning the exposures 
do not affect the probabilities of respnoses), to 1 
(or 100\%) (meaning that under different exposures, workers provide completely
different sets of responses).

@@

Using a na\"ive Bayes classifier (\# in the 
supplementary material we discuss the choice of classifier, the
approach to measuring accuracy, and draw a comparison to the results obtained 
if a support vector machine is used), we found that intertask effects 
lead to a 30\% bias between workers from the \textit{food} and 
\textit{objects} treatments.

There has been considerable interest in the effects that framing can have
on microtask responses
\cite{Kinnaird2012281,chandler2013breaking,thibodeau2013natural}, so,
as a point of comparison, we performed a similar experiment 
in which we framed the purpose of the tasks.  In the experiment 
\textit{frame-food-objects}, workers were either told that 
the tasks were ``Funded by the laboratory for the visual perception of Food 
and Ingredients'', or ``\ldots of Objects and Tools''.  
Framing did induce statistically significant changes in the frequencies of 
word usage (as determined by $\chi^2$ test; $p=0.0012$), albeit at a much 
lower significance than for intertask effects.  But we did not find a 
significant amount of framing-induced bias, using a na\"ive Bayes classifier 
(see Fig.~\ref{fig:theta}).  This suggests that prior task exposure is a 
stronger priming modality than framing.

We conducted variants of these experiments using different images, to see 
whether this trend was robust.  In the experiment 
\textit{intertask-food-culture},
workers were either assigned to a \textit{food} or \textit{culture} treatment.
As before, the initial tasks for the \textit{food} treatment contained images 
depicting food (\#) (see Fig.~\ref{fig:task}D). 
In the \textit{culture} treatment, 
the initial tasks contained images depicting cultural scenes 
(of dance, sport, or music) (see Fig.~\ref{fig:task}E).  The test tasks, which were identical for both
treatments, included images depicting meals of identifiable cultural 
origin 
(see Fig.~\ref{fig:task}F).  Results for this experiment again showed a 
strong bias as a result of intertask effects (about 50\%) 
(see Fig.~\ref{fig:theta}A).  Again, we compared intertask effects to 
framing, by performing an experiment (\textit{frame-food-culture}) using 
the same test tasks as in \textit{intertask-food-culture}, in which we exposed
workers to a statement framing the purpose of the tasks as the recognition 
of either food or culture.  
For this framing experiment, we did not observe a statistically significant 
difference between the labels attributed by the \textit{food} and 
\textit{culture} treatments during test tasks (see Fig.~\ref{fig:theta}).

@@

To deepen our understanding, we investigated workers' lexical richness in 
reference to food, that is, the number of \textit{unique} food-related words
used.  Even if workers provide an abundance of food-related words, if the 
same words are often repeated, there will of course be a smaller total 
diversity.  This could happen, for example, if workers use generic references 
to food.
Both \textit{intertask} experiments showed that food-exposed workers had 
greater lexical richness, in reference to food, than their non-food-primed 
counterparts (as much as 20\% more) (see Fig.~\ref{fig:specificity}).  
The increase in lexical richness in the case of 
\textit{intertask-food-culture} is quite interesting, because in that 
experiment, food-primed workers made fewer total references to food.  
We also observed enrichment of the food-related lexicon in the 
\textit{priming-food-objects}, and \textit{echo-food-objects} experiments, 
although to a lesser extent.

The observations regarding lexical richness lead us to suspect that 
initial tasks might influence workers to use more refined or specialized 
words, when refering to aspects of content that had been present in the 
initial tasks (i.e. food).  This would help explain why, in the case of 
\textit{intertask-food-culture}, we observed a significantly greater 
\textit{diversity} of food-related words despite their significantly lesser 
\textit{total amount}.
To test whether exposure to food in initial tasks increased the specialization
of food-references, we used the wordnet corpus to operationalize the 
notion of word specialization.  Wordnet provides a set of hypernym-hyponym 
relations between \# English nouns.  Hypernyms are generalizations 
(for example, ``bread'' is a hypernym for 
``pumpernickel''), while hyponyms are specializations.
For each experiment, we compared food-related words provided by workers from 
one treatment to those from the other, and tallied the excess number of cases
where one treatment's words were more specialized than the other.
Expressed as a percentage of their total, this tally served to 
measure the relative specialization of the vocabulary used in one 
treatment's vocabulary relative to the other 
(\# full detail for this calculation is provided the supplementary text) 
(see Fig.~\ref{fig:specificity}).

In all experiments except \textit{frame-food-culture}, food-primed workers 
used significantly more specialized words than their non-food-primed 
counterparts (about 15\% more; see Fig.~\ref{fig:specificity}C).
It is interesting that such substantial increases in both the lexical 
richness and specialization of food-related words 
held for \textit{intertask-food-culture}, where, as mentioned, we observed 
that food-exposed workers had made \textit{fewer} references to food overall. 
These observations point 
to countervailing factors: one factor tending to activate the more 
specialized and less common food-related words 
(yielding greater lexical richness and specialization), and the other tending 
to suppress certain, presumably more common and generic words 
(yielding fewer food-related words in total).

@@

Taken together, our results can be 
explained through a combination of positive and negative priming effects.
Positive priming (usually simply ``priming'') occurs when a prior stimulus 
predisposes a person to give certain responses in an ensuing task, and
is often observed as an increase in the speed or accuracy of a response, or
the ability to recognize features of a briefer, quieter, or noisier stimulus 
(\#).
Negative priming occurs when, after being repeatedly exposed to a stimulus 
considered to be non-salient, a person begins to ignore that 
stimulus (\#).

@@

Prior to this study, little thought appears to have been given to the 
ordering and bundling of microtasks.  But our findings show that these are 
powerful design factors.  A natural response to our findings 
might be simply to randomize task ordering, a practice that is commonly 
employed.  But the sheer extent of bias that we observed leads us to question 
whether that is the best approach.  Although the bias is random in 
that case, it is still on the order of 30-50\%, which is a major contribution
to noise in the results.  It would be preferable to understand intertask
effects in greater depth so that they might \textit{controlled} rather than
simply randomized.

Our findings suggest that, through careful task engineering,
it might be possible to achieve greater quality and reproducibility
of responses.  
A consistent goal in human computation is the achievement of expert-level
judgments from non-expert workers (\#).  This has been achieved in some
applications (\#). The distinction between experts and novices can be 
attributed in part to the posession of specialized knowledge and heuristics. 
But to a large degree, this distinction lies in the ability of 
experts to direct their focus toward salient features, while filtering out 
non-salient ones \cite{kellman2009perceptual}.  Using strategic exposure to
prior tasks, it might be possible to guide workers' focus and salience 
attribution, enabling expert-level judgment in a wider variety of 
crowdsourced applications.
We anticipate that future work will yield techniques to control intertask 
effects, enabling one to both reduce unwanted bias, as well as tune the 
focus, diversity, and specificity of worker responses.

Here we have shown that a large bias can be introduced into microtask 
responses, simply from the performance of earlier tasks. 
These effects are much stronger than framing.  
While our findings raise serious concerns about the 
current state of microtask design, they also reveal potential opportunities 
for more 
refined control over worker focus and acuity.  Any application that relies
on the repetitive use of human judgment is likely subject to the phenomena
we described here.  Exploring the concerns and opportunities posed by 
intertask effects are important directions for future work, 
in the effort to develop performant and reliable methods for human 
computation.
\bibliography{newbib}
\bibliographystyle{Science}
